pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-05-07 bmc_article.tex
Updates from Overleaf

Here, we develop a flexible R framework for pipeline comparison and apply it to the evalution of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 3). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}, see Methods for more details). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 10). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out. This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) often leads to high clustering performance (e.g. simMix and Zhengmix4 datasets) and always retain most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.  Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely (e.g. Zheng datasets). On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 11). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:norm}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets. It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth. \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 13) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{SCTransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 14). We also confirmed that, especially in the UMI-based datasets, \textit{SCTransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 15), at little apparent cost on computation time (Supplementary Figure 16).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 17A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 17B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 18-19). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 18-20). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 18-19). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 20).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}). Again, the clustering on the biggest datasets (simMix and Zhengmix) was more challenging for most methods. Several of them could not identify the true number of clusters in the simMix1 and Zhengmix8eq datasets, which have the highest number of subpopulations, or lead to very low mean silhouette width and accuracy. Overall however, the clustering accuracy was higher for methods that include weighting or scaling. Among them, GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 13). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 14 and 21). This tended to affect the accuracy of the clustering (Supplementary Figure 22), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). Figure \ref{fig:dim}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:dim}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 23) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

>>>>

Here, we develop a flexible R framework for pipeline comparison and apply it to the evalution of the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Supplementary Figure 1 for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification; 2) cell filtering; 3) normalization; 4) feature selection; 5) dimension reduction; and, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide practical recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 4). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 5 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 6), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 7). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to six methods with varying stringency; three methods based on deviations to MADs with increasing levels of stringency (named \textit{lenient}, \textit{default}, \textit{stringent}, \textit{very stringent}) and two methods based on \textit{scater}'s \textit{runPCA} using all (\textit{pca.all}) or selected covariates (\textit{pca.sel}, see Methods for more details). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:filtering}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:filtering}B and Supplementary Figure 11). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious. Most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:filtering}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) rather than on whole dataset resulted in nearly no cells being filtered out (Supplementary Figure 12). This suggests that stringent filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. On the opposite, too lenient filtering (or lack thereof) can lead to misclassified cells, as was for instance the case especially with the Zhengmix4 datsets (Figure \ref{fig:filtering}B). We found that a relatively mild filtering (`default' - see methods) often leads to high clustering performance (e.g. simMix and Zhengmix4 datasets) and always retain most of the cells in each subpopulation (e.g. Koh and Zhengmix8eq). This suggests that a mild ('default') filtering is a very good trade-off between clustering accuracy and subpopulations preservation.  Finally, the doublet removal step in conjunction with filtering  resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy not only in datasets that do include heterotypic doublets (mixology datasets), but also some others in which such doublets are unlikely (e.g. Zheng datasets). On the Smart-Seq and simulated datasets however, the removal of doublets had little to no impact on any of the two measures. Overall, we therefore recommend the use of doublet removal followed by our `default' filtering (see Methods) or a similar approach.

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 13). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:norm}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. In view of the subpopulation silhouette and ARI at true number of clusters, it seems that all methods perform worse on the three Zhengmix datasets, that are the datasets with the highest number of cells and the smaller feature detection rate. Scaling (e.g. `seurat (no scale)' vs `seurat') tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave equal  or poorer performance on MI, ARI and ARI at true number of clusters, with a notable negative impact when \textit{Seurat}'s regressing was applied to the Koh and Kumar datasets. It is possible that the difference between \textit{Seurat} and \textit{SCTransform}'s regression originates from the use of Pearson's residuals performed during the second round regression of \textit{SCTransform}, which are then independent of sequencing depth. \textit{SCTransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).\textit{scVI} (Supplementary Figure 15) performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. For time and resource cost reasons, \textit{scVI} was run on a reduced set of pipelines and compared to \textit{SCTransform} (more details in the Methods section).

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{SCTransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 16). We also confirmed that, especially in the UMI-based datasets, \textit{SCTransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 17), at little apparent cost on computation time (Supplementary Figure 18).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{SCTransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{SCTransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:selection}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The ARI at true number of clusters was near perfect (>0.94) among the smallest datasets but among bigger datasets (simMix and ZhengMix) the performance of the methods was lower and more variable (Figure \ref{fig:selection}B). Non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 22).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Supplementary Figure 23). Again, the clustering on the biggest datasets (simMix and Zhengmix) was more challenging for most methods. Several of them could not identify the true number of clusters in the simMix1 and Zhengmix8eq datasets, which have the highest number of subpopulations, or lead to very low mean silhouette width and accuracy. Overall however, the clustering accuracy was higher for methods that include weighting or scaling. Among them, GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts. Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 15). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 16 and 24). This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

We compared the various estimates of dimensionality in their ability to retrieve the intrinsic number of dimensions in a dataset, based on Seurat's weighted PCA space. As a first approximation of the true dimensionality, we computed the variance in each principal component that was explained by the subpopulations, which sharply decreased after the first few components in most datasets (Figure \ref{fig:dim}A). Figure \ref{fig:dim}B shows the difference between the dimension estimates of the above methods and that based on the subpopulations (i.e. from Figure \ref{fig:dim}A). Of note, the methods differ widely in view of their computing time (Figure \ref{fig:dim}B) and we saw no relationship between the accuracy of the estimates and the complexity of the method. Reasoning that over-estimating the number of dimensions is less problematic than under-estimating it, we kept the former methods for a full analysis of their impact on clustering (Figure \ref{fig:dim}C-D), when combined with \textit{SCTransform} or the standard \textit{Seurat} normalization. Although most methods performed well on the various clustering measures, the global maximum likelihood based on translated Poisson Mixture Model (\textit{maxLikGlobal}, using the 20 nearest neighbors, although estimates were relatively stable across decent values of \textit{k}; see Supplementary Figure 26) provided the dimensionality estimate that best separated the subpopulations (Figure \ref{fig:dim}C) and resulted in the best clustering accuracy (Figure \ref{fig:dim}D). This method systematically selected many more components than were associated with the subpopulations, suggesting that although these additional components appear individually uninformative, in combination they nevertheless contribute to classification. An implication is that the commonly-used elbow method most likely underestimates dimensionality.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figures 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 27), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.
